---
title: "MIDTERM DRAFT"
author: "Julia Haas"
date: "2023-11-02"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE}
library(ggplot2)
hd.data<-read.csv("processed.heartdisease.txt", na.strings="?")
attach(hd.data)
```

1. Are there notable associations/relationships between some of the variables? Are there any meaningful groups of variables that exhibit these associations? If so, describe them.

 
```{r, echo=FALSE, warning=FALSE}
library(stats)
library(cluster)
library(MVA)
quant<-hd.data[, c("age", "trestbps", "chol", "thalach", "oldpeak")]
round(cor(quant),2)    #correlation matrix

oldpeak.thalach<-table(hd.data[,c(10,13)])
chisq.test(oldpeak.thalach)

#thal and diagnosis
thal.num<-table(hd.data[,c(13,14)])
thal.num
chisq.test(thal.num)
```

2. Is there a way to graphically represent the raw data for the 303 patients and draw conclusions about the data set from such a graph?
graphs, visualizing the data? 


```{r, echo=FALSE, warning=FALSE}
library(ggbiplot)
hd.data.scaled<-(scale(quant))
hd.data.pc<- princomp(hd.data.scaled, cor=T)   #pca of quantitative data
hd.data.pc.sum <- summary(hd.data.pc, loadings=T)
hd.data.pc.sum$loadings  #loadings and components
summary(hd.data.pc)
num<-factor(num)

qplot(1:(length(hd.data.pc$sdev)),  (hd.data.pc$sdev)^2, geom='line', main="Scree Plot", xlab="Number of Components", ylab="Eigenvalue Size", xlim=c(1,5)) #scree plot

ggbiplot(hd.data.pc, obs.scale = 0.1, var.scale = 0.3, labels.size = 0.1, varname.size = 3, var.axes = TRUE, groups = num, ellipse=FALSE) + scale_color_brewer()+theme(panel.background = element_rect(fill="grey80"))
```

3. What are the basic underlying groups that the individuals form? Can you plot the data in a small number of dimensions, showing the group separation of the patients?


```{r, echo=FALSE, warning=FALSE}
library(MASS)
library(ggbiplot)


dist.all<-as.matrix(daisy(hd.data, metric="gower")) #distance matrix including categorical data
mds<-isoMDS(dist.all, k = 5)   #mds including categorical data
sex<-factor(sex)
num<-factor(num)
plot(mds$points, type = "n", xlab="Coordinate 1", ylab="Coordinate 2", main="Non-metric  MDS")
text(mds$points, labels = rownames(mds$points), cex=.7)
mds.xy <- data.frame(mds$points)

ggplot(mds.xy, aes(X1,X2, color = num)) +geom_point((aes(shape=sex)))+  scale_size(range = c(1.5, 2))
ggplot(mds.xy, aes(X2,X3, color = num)) +geom_point((aes(shape=sex)))+  scale_size(range = c(1.5, 2))
```

```{r, fig.width=10, warning=FALSE}
library(factoextra)
library(dendextend)
#clustering

clustframe<-hd.data
clust1<-hclust(daisy(clustframe, metric="gower"))
clust1$labels<-clustframe$num
cut <- cutree(clust1, k = 4)
plot(clust1)
rect.hclust(clust1 , k = 4, border = 4:6)

```

4. Are there interesting differences in any of the recorded fields with respect to heart disease diagnosis? 



5. If the researchers were to investigate a new patient observation that had known measurements for the 13 explanatory variables, could we determine a rule for predicting that patient's heart disease status (no heart disease vs presence of heart disease)? How accurate could you expect such a rule to be?

```{r, echo=FALSE, warning=FALSE}
library(MASS)
library(MVN) 

mvn(quant, mvnTest="hz")
mvn(quant, mvnTest="mardia")
mvn(quant, mvnTest="royston")
mvn(quant, mvnTest="hz", multivariatePlot="qq")


heartlda <- lda(num ~ age + trestbps + chol + thalach + oldpeak, data=hd.data)
heartlda

correct<-rep(0,times=nrow(hd.data) )
for (j in 1:nrow(hd.data) ) {
  mydis<-lda(grouping=hd.data$num[-j], x=hd.data[-j,1:3], prior=c(0.54125413, 0.18151815, 0.11881188, 0.11551155, 0.04290429))
  mypred<-predict(mydis,newdata=hd.data[j,1:3])$class
  correct[j] <- (mypred==hd.data$num[j])
}
cv.misclass <- 1-mean(correct)
cv.misclass

```

6. In particular, we have a new patient who is a 60 year old female. Her symptoms are non-anginal pain, a resting blood pressure of 102 mm Hg, a cholesterol measurement of 318 mg/dl, low fasting blood sugar, normal resting electrocardiographic results, a maximum heart rate of 160 beats/minute, no exercise-induced angina, no ST depression induced by exercise relative to rest, upsloping peak ST segment, only 1 colored major vessel, and normal thal diagnosis. Would you predict this patient to have heart disease? How confident are you in the classification?


```{r, echo=FALSE}
newobs <- rbind( c(60,102,318, 160, 0) )
dimnames(newobs) <- list(NULL,c("age", "trestbps", "chol", "thalach", "oldpeak"))
newobs <- data.frame(newobs)
predict(heartlda,newdata=newobs)$class

predict(heartlda,newdata=newobs)$posterior

zero.df <- hd.data[num == "0", ] 
nonzero.df <- hd.data[num != "0", ] 
#------------------------------------------------------------------------
cp.table <- table(zero.df$cp) 
ratio3 <- prop.table(cp.table)  #ratio of cp rates no heart disease
ratio3    

cp.table2 <- table(nonzero.df$cp)  
ratio4 <- prop.table(cp.table2)  #ratio of cp rates with heart disease
ratio4
# The majority of patients who did not have heart disease had non-anginal pain, 41.46%, unlike the majority of patients who had some form of heart disease, most showed asymptomatic. 
#------------------------------------------------------------------------
thal.table <- table(zero.df$thal) 
ratio1 <- prop.table(thal.table)  #ratio of thal rates no heart disease
ratio1    

thal.table2 <- table(nonzero.df$thal)  
ratio2 <- prop.table(thal.table2)  #ratio of thal rates with heart disease
ratio2
# The majority of patients who did not have heart disease had a normal thal diagnosis, 79.14%. The majority of patients who had heart disease had a reversible defect diagnosis.
#------------------------------------------------------------------------

restecg.table <- table(zero.df$restecg) 
ratio5 <- prop.table(restecg.table)  #ratio of restecg rates no heart disease
ratio5    

restecg.table2 <- table(nonzero.df$restecg)  
ratio6 <- prop.table(restecg.table2)  #ratio of restecg rates with heart disease
ratio6
# The majority of patients who did not have heart disease had normal resting electroencephalographic results, 57.927%. The majority of heard disease patients had results showing probable or definite left ventricular hypertrophy by Estes' criteria.
#------------------------------------------------------------------------
ca.table <- table(zero.df$ca) 
ratio7 <- prop.table(ca.table)  #ratio of ca rates no heart disease
ratio7    

ca.table2 <- table(nonzero.df$ca)  
ratio8 <- prop.table(ca.table2)  #ratio of ca rates with heart disease
ratio8
# The majority of non-heart disease patients had 0 colored vessels, the majority of heart disease patients also had 0 colored vessels, 33.33% with 1 colored vessel a close second at 31.88%. This could be meaningful. 
#------------------------------------------------------------------------
fbs.table <- table(zero.df$fbs) 
ratio9 <- prop.table(fbs.table)  #ratio of fbs rates no heart disease
ratio9    

fbs.table2 <- table(nonzero.df$fbs)  
ratio10 <- prop.table(fbs.table2)  #ratio of fbs rates with heart disease
ratio10
# majority of non-heart disease patients had low fasting blood sugar, 85.98%. That being said, the majority of heart disease patients also had low fasting blood sugar,84.17%. 
#------------------------------------------------------------------------
exang.table <- table(zero.df$exang) 
ratio11 <- prop.table(exang.table)  #ratio of exang rates no heart disease
ratio11   

exang.table2 <- table(nonzero.df$exang)  
ratio12 <- prop.table(exang.table2)  #ratio of exang rates with heart disease
ratio12

#The majority of non-heart disease patients had no exercise induced angina, 85.98% while the majority of heart disease patients had exercise induced angina, 54.67%. 
#------------------------------------------------------------------------
slope.table <- table(zero.df$slope) 
ratio13 <- prop.table(slope.table)  #ratio of slope rates no heart disease
ratio13   

slope.table2 <- table(nonzero.df$slope)  
ratio14 <- prop.table(slope.table2)  #ratio of exang rates with heart disease
ratio14
# The majority of non-heart disease patients had an upsloping peak ST segment, 64.34% unlike the majority of heart disease patients who had a majority of flat peak ST segment, 64.47%. 
#------------------------------------------------------------------------
sex.table <- table(zero.df$sex) 
ratio15 <- prop.table(sex.table)  #ratio of women vs men with no heart disease
ratio15    

sex.table2 <- table(nonzero.df$sex)  
ratio16 <- prop.table(sex.table2)  #ratio of women vs men with heart disease
ratio16

ratio.sex<-prop.table(table(sex))
ratio.sex

# In the dataset, 32% of the patients were female and 68% of the patients were male. Of those patients, 82% of patients diagnosed with heart disease were male and 18% were female. 
```
